\pkg{DNest4}: Diffusive Nested Sampling in \proglang{C++} and \proglang{Python}

### 2018-08-20 paper/ms.tex
Response

\begin{table}[t!]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
\hline
Package     &   Easy to     & High   &   Multimodal     & Dependent   &   Phase   &   Computes $Z$?\\
            &   implement & dimensions? & distributions? & distributions? & changes? & \\
& models? & & & & &\\
\hline
\pkg{DNest4}    & \xmark  & \cmark & \cmark & \cmark &\cmark &\cmark\\
\pkg{emcee}    & \cmark  & \xmark & \xmark & \cmark &\xmark &\xmark\\
\pkg{JAGS}    & \cmark  & \cmark & \xmark & \xmark & \xmark &\xmark\\
\pkg{MultiNest}    & \cmark  & \xmark & \cmark & \cmark & \cmark &\cmark\\
\pkg{Stan}    & \cmark  & \cmark & \xmark & \cmark & \xmark &\xmark\\
\hline
\end{tabular}}
\caption{A simplified summary of the advantages and disadvantages of some
Bayesian computation software packages.\label{tab:packages}}
\end{table}
\newpage
\section{Markov chain Monte Carlo}\label{sec:mcmc}
DNS is build upon the Metropolis-Hastings algorithm.
In this algorithm, the acceptance probability $\alpha$
is given by
\begin{align*}
\alpha &= \min\left(1,
\frac{q(\params'|\params)}{q(\params | \params')}
\frac{\pi(\params')}{\pi(\params)}\frac{L(\params')}{L(\params)}
\right),
\end{align*}
where $q(\theta' | \theta)$ is the proposal distribution used to generate
a new position $\theta'$ from the current position $\theta$. Often,
$q$ is {\em symmetric} so that the $q$ terms cancel in the acceptance probability.

@@

\section{Postprocessing}\label{sec:postprocessing}
The output files themselves are typically not immediately useful.
The goal of running
nested sampling is usually to obtain posterior samples and the marginal likelihood $Z$,
whereas \code{sample.txt} only contains samples from the mixture of constrained
priors. Additional
post-processing is required. This can be achieved by running the following
\proglang{Python} function\footnote{Alternatively, the file
\code{showresults.py} in the example directory runs this function, and then
calls the code in \code{display.py} to create a further plot shown in
Figure~\ref{fig:fig3} on the right.}:
%
\begin{CodeChunk}
\begin{CodeInput}
>>> import dnest4
>>> dnest4.postprocess()
\end{CodeInput}
\end{CodeChunk}
%
This produces the three diagnostic plots in Figures~\ref{fig:fig1}
and~\ref{fig:fig3} on the left, along with the following output:
%
\begin{CodeChunk}
\begin{CodeOutput}
log(Z) = -175.47872682863076
Information = 15.166340875399044 nats.
Effective sample size = 1251.3576752091165
\end{CodeOutput}
\end{CodeChunk}
%
These are the natural log of the marginal likelihood, the
information
\begin{align*}
\mathcal{H} &= \int p(\theta|D, M)
\ln\left[\frac{p(\theta | D, M)}{p(\theta | M)}\right] \, d\theta,
\end{align*}
which quantifies the degree to which $D$ restricted the
range of possible $\theta$ values,
and the effective sample size,
or number of saved particles with significant posterior weight.
The \code{postprocess} function also saves a file,
\code{posterior\_sample.txt}, containing posterior samples (one per row).
Unfortunately, it is harder to compute justified error bars on $\ln(Z)$
in DNS than it is in standard nested sampling.
%The \code{postprocess} function does include some options
%for calculating error bars, but they are not used by default,
%and we do not describe them here.
%Hence, the zero-sized error bars on $\ln(Z)$ and $\mathcal{H}$ should
%be ignored.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.4\textwidth]{rjobject_pgm.pdf}
\caption{A PGM showing the kind of prior information the `\code{RJObject}'
template class expresses.
Figure created using Daft (\url{http://daft-pgm.org}).\label{fig:rjobject_pgm}}
\end{figure}

@@

We now describe the specific prior distributions we used in this example.
The prior for $N$ was
\begin{align*}
p(N) \propto \frac{1}{N+1}
\end{align*}
for $N \in \{1, 2, \ldots, 100\}$.
For the conditional prior of the Gaussian parameters,
we used Laplace (biexponential)
distributions\footnote{A Laplace
distribution with location parameter $a$ and scale parameter
$b$ has density $p(x | a, b) = \frac{1}{2}\exp\left(-\frac{1}{b}|x - a|\right)$.}.
Normal distributions would be more conventional, but the analytic cumulative
distribution function (CDF) of the Laplace distribution makes it easier to
implement. These were:
\begin{align}
\mu_j &\sim \textnormal{Laplace}(a_\mu, b_\mu)\label{eqn:cond_prior1}\\
\ln\sigma_j &\sim \textnormal{Laplace}(a_{\ln\sigma}, b_{\ln\sigma})\label{eqn:cond_prior2}\\
\ln W_j &\sim \textnormal{Laplace}(0, b_{\ln W}).\label{eqn:cond_prior3}
\end{align}
The location parameters are denoted with $a$ and scale parameters with $b$.
These priors express the idea that the centers, widths, and relative weights
of the Gaussian mixture components are probably clustered around some typical
value.

@@

A simple ``rejection sampling'' version of
ABC works by sampling the joint prior distribution for the parameters and
data
\begin{align*}
p(\theta, D | M) &= p(\theta | M)p(D | \theta, M)
\end{align*}
and rejecting samples for which $D \neq D_{\rm observed}$, so that the
samples represent
\begin{align}
p(\theta, D | D = D_{\rm observed}, M) &\propto p(\theta | M)p(D | \theta, M)
\mathds{1}\left[D = D_{\rm observed}\right].\label{eqn:abc_posterior}
\end{align}
The marginal distribution for $\theta$ is also the conditional distribution
$p(\theta | D = D_{\rm observed}, M)$, that is,
the posterior\footnote{Incidentally, this usage of the {\em joint} posterior
distribution for the parameters {\em and the data} provides a bridge
connecting Bayesian updating to maximum entropy updating
\citep{caticha2006updating,giffin2007updating}.}.

@@

\subsection{ABC example}
A simple ABC example is included in the directory
\code{code/Examples/ABC}. The example tries to infer the mean
$\mu$ and standard deviation $\sigma$ of a normal distribution
from a set of samples. The sampling distribution is
\begin{align*}
x_i &\sim \textnormal{Normal}(\mu, \sigma^2).
\end{align*}
and the prior is
\begin{align*}
\mu &\sim \textnormal{Uniform}(-10, 10),\\
\ln\sigma &\sim \textnormal{Uniform}(-10, 10).
\end{align*}
However, instead of conditioning on the full dataset $\{x_i\}$, the example
uses the minimum and maximum values in the data,
$\min(\{x_i\})$ and $\max(\{x_i\})$, as summary statistics\footnote{This is
not a serious application of ABC, since the likelihood function can easily
be evaluated. It is included to demonstrate the principle of using
nested sampling as an ABC method.}.

>>>>

\begin{table}[t!]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
\hline
Package     &   Easy to     & High   &   Multimodal     & Dependent   &   Phase   &   Computes $Z$?\\
            &   implement & dimensions? & distributions? & distributions? & changes? & \\
& models? & & & & &\\
\hline
\pkg{DNest4}    & \xmark  & \cmark & \cmark & \cmark &\cmark &\cmark\\
\pkg{emcee}    & \cmark  & \xmark & \xmark & \cmark &\xmark &\xmark\\
\pkg{JAGS}    & \cmark  & \cmark & \xmark & \xmark & \xmark &\xmark\\
\pkg{MultiNest}    & \cmark  & \xmark & \cmark & \cmark & \cmark &\cmark\\
\pkg{Stan}    & \cmark  & \cmark & \xmark & \cmark & \xmark &\xmark\\
\hline
\end{tabular}}
\caption{A simplified summary of the advantages and disadvantages of some
Bayesian computation software packages.\label{tab:packages}}
\end{table}
\newpage
\section{Markov chain monte carlo}\label{sec:mcmc}
DNS is build upon the Metropolis-Hastings algorithm.
In this algorithm, the acceptance probability $\alpha$
is given by
\begin{align*}
\alpha &= \min\left(1,
\frac{q(\params'|\params)}{q(\params | \params')}
\frac{\pi(\params')}{\pi(\params)}\frac{L(\params')}{L(\params)}
\right),
\end{align*}
where $q(\theta' | \theta)$ is the proposal distribution used to generate
a new position $\theta'$ from the current position $\theta$. Often,
$q$ is {\em symmetric} so that the $q$ terms cancel in the acceptance probability.

@@

\section{Postprocessing}\label{sec:postprocessing}
The output files themselves are typically not immediately useful.
The goal of running
nested sampling is usually to obtain posterior samples and the marginal likelihood $Z$,
whereas \code{sample.txt} only contains samples from the mixture of constrained
priors. Additional
post-processing is required. This can be achieved by running the following
\proglang{Python} function:
%
\begin{CodeChunk}
\begin{CodeInput}
>>> import dnest4
>>> dnest4.postprocess()
\end{CodeInput}
\end{CodeChunk}

Alternatively, the file
\code{showresults.py} in the example directory runs this function, and then
calls the code in \code{display.py} to create a further plot shown in
Figure~\ref{fig:fig3} on the right.

%
This produces the three diagnostic plots in Figures~\ref{fig:fig1}
and~\ref{fig:fig3} on the left, along with the following output:
%
\begin{CodeChunk}
\begin{CodeOutput}
log(Z) = -175.47872682863076
Information = 15.166340875399044 nats.
Effective sample size = 1251.3576752091165
\end{CodeOutput}
\end{CodeChunk}
%
These are the natural log of the marginal likelihood, the
information
\begin{align*}
\mathcal{H} &= \int p(\theta|D, M)
\ln\left[\frac{p(\theta | D, M)}{p(\theta | M)}\right] \, d\theta,
\end{align*}
which quantifies the degree to which $D$ restricted the
range of possible $\theta$ values,
and the effective sample size,
or number of saved particles with significant posterior weight.
The \code{postprocess} function also saves a file,
\code{posterior\_sample.txt}, containing posterior samples (one per row).
Unfortunately, it is harder to compute justified error bars on $\ln(Z)$
in DNS than it is in standard nested sampling.
%The \code{postprocess} function does include some options
%for calculating error bars, but they are not used by default,
%and we do not describe them here.
%Hence, the zero-sized error bars on $\ln(Z)$ and $\mathcal{H}$ should
%be ignored.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.4\textwidth]{rjobject_pgm.pdf}
\caption{A PGM showing the kind of prior information the `\code{RJObject}'
template class expresses.
Figure created using \pkg{Daft} \citep{daft}.\label{fig:rjobject_pgm}}
\end{figure}

@@

We now describe the specific prior distributions we used in this example.
The prior for $N$ was
\begin{align*}
p(N) \propto \frac{1}{N+1}
\end{align*}
for $N \in \{1, 2, \ldots, 100\}$.
For the conditional prior of the Gaussian parameters,
we used Laplace (biexponential)
distributions. A Laplace
distribution with location parameter $a$ and scale parameter
$b$ has density $p(x | a, b) = \frac{1}{2}\exp\left(-\frac{1}{b}|x - a|\right)$.
Normal distributions would be more conventional, but the analytic cumulative
distribution function (CDF) of the Laplace distribution makes it easier to
implement. These were:
\begin{align}
\mu_j &\sim \textnormal{Laplace}(a_\mu, b_\mu)\label{eqn:cond_prior1}\\
\ln\sigma_j &\sim \textnormal{Laplace}(a_{\ln\sigma}, b_{\ln\sigma})\label{eqn:cond_prior2}\\
\ln W_j &\sim \textnormal{Laplace}(0, b_{\ln W}).\label{eqn:cond_prior3}
\end{align}
The location parameters are denoted with $a$ and scale parameters with $b$.
These priors express the idea that the centers, widths, and relative weights
of the Gaussian mixture components are probably clustered around some typical
value.

@@

A simple ``rejection sampling'' version of
ABC works by sampling the joint prior distribution for the parameters and
data
\begin{align*}
p(\theta, D | M) &= p(\theta | M)p(D | \theta, M)
\end{align*}
and rejecting samples for which $D \neq D_{\rm observed}$, so that the
samples represent
\begin{align}
p(\theta, D | D = D_{\rm observed}, M) &\propto p(\theta | M)p(D | \theta, M)
\mathds{1}\left[D = D_{\rm observed}\right].\label{eqn:abc_posterior}
\end{align}
The marginal distribution for $\theta$ is also the conditional distribution
$p(\theta | D = D_{\rm observed}, M)$, that is,
the posterior. Incidentally, this usage of the {\em joint} posterior
distribution for the parameters {\em and the data} provides a bridge
connecting Bayesian updating to maximum entropy updating
\citep{caticha2006updating,giffin2007updating}.

@@

\subsection{ABC example}
A simple ABC example is included in the directory
\code{code/Examples/ABC}. The example tries to infer the mean
$\mu$ and standard deviation $\sigma$ of a normal distribution
from a set of samples. The sampling distribution is
\begin{align*}
x_i &\sim \textnormal{Normal}(\mu, \sigma^2).
\end{align*}
and the prior is
\begin{align*}
\mu &\sim \textnormal{Uniform}(-10, 10),\\
\ln\sigma &\sim \textnormal{Uniform}(-10, 10).
\end{align*}
However, instead of conditioning on the full dataset $\{x_i\}$, the example
uses the minimum and maximum values in the data,
$\min(\{x_i\})$ and $\max(\{x_i\})$, as summary statistics. This is
not a serious application of ABC, since the likelihood function can easily
be evaluated. It is included to demonstrate the principle of using
nested sampling as an ABC method.
